When Do You Repeat Yourself?\\ Voices from the Trenches of Linux Kernel Maintainers on Code Duplication

### 2025-10-28 papers/techdebt2026/techdebt.tex
techdebt2026: techdebt: add mentoring of students description in participant-as-observer
We were not citing how we supported students in their patch submissions
to the Linux kernel, which omitted a great deal of how the second phase
can be considered a participant-as-observer observation. So, add it to
the corresponding subsection, making adjustments for fluidity.

Signed-off-by: David Tadokoro <davidbtadokoro@usp.br>

As the first option, we ran the tool on the IIO subsystem, creating a list of duplicated function pairs. Then, we curated the list even more to get more actionable duplications for the students using this approach:

\begin{enumerate}
    \item We filtered the list to keep only duplicated pairs that happened in the same file, because the IIO subsystem is very decentralized, and a duplication between two different drivers may involve the cooperation of different maintainers and/or the unification of totally different interfaces;
    \item We removed duplications that were too short and would not represent a significant contribution;
    \item We manually ranked code duplications and marked some duplications with insights based on personal opinion and perceptions over the remaining entries. This was done to guide students so that they could approach the technical debt with a greater chance of having their contribution merged into the code.
\end{enumerate}

The students' task was to select a recommended entry from the list, refactor the code to eliminate these duplications, and then submit their patches to the IIO mailing list for the maintainers' review.
%TODO: explicar o processo do envio para o mentores/monitores, o tipo de feedback que recebiam antes enviarem.

For the second alternative, students were offered an experience similar to our initial complete participant observation. In this approach, students were responsible for the entire workflow: executing the tool, independently analyzing the results to identify a potential deduplication, creating a patch, and sending the patch to the driver's maintainers.
%TODO: aqui, no caso do Sebastian, também houve a revisão intermediária dos mentores?

@@

Using blog posts, the groups documented their experiences and approaches to refactoring the duplications, as well as their experiences submitting a patch to the Linux kernel~[omitted]. We also analyzed these posts to understand the refactoring patterns used in deduplications, the experience of sending patches, and the opinions of subsystem maintainers on them.
%TODO: é fácil e queremos disponibilizar esses blogs também no pacote de replicação?

@@

Since we suffered a delay in the review process, we directly contacted a maintainer to understand why. From this conversation, we understand that one reason for the delay was that duplicated code enhances the independence of GPU driver code. This approach allows developers to make changes to a specific GPU without needing to test compatibility with others, which helps save a significant amount of time and effort. This example of hardware variations, also mentioned by Kapser and Godfrey~\cite{cloneharm}, leads us to the primary conclusion of our complete participant observation:
%
\begin{highlight-box}{Takeaway T1}
Driver developers do not necessarily view duplicated code negatively, which contradicts standard software engineering practices.
\end{highlight-box}

@@

By the end of the feedback cycle, the patch from Group 4 was accepted, but the final implementation differed substantially from the original without using macros. The change was primarily driven by a maintainer's feedback: \textit{``I think the old code is more readable than hiding the values in a macro even if it is duplicating a few lines of code.''}.
%
This response category (also from groups 1, 2, and 10) revealed a recurring concern among maintainers that went beyond functional correctness or redundancy reduction. Instead, it reflected a preference for code clarity and comprehension. Building on these consistent insights from multiple review interactions, we formulated our second takeaway:
%\vspace{-.5cm}
\begin{highlight-box}{Takeaway T2 (Readability)}
Maintainers prioritized readability over removing duplications in many different contexts. In their view, going against the \textbf{DRY principle} was worth it if it resulted in easier-to-read and understand code.
\end{highlight-box}

Still focusing on Group 1, maintainers considered the patch irrelevant. The reason was that Group 1 chose to apply the refactoring to a device containing a register that was both read-only and non-volatile. Thus, the technical debt selected by Group 1 was not merely a duplication but an incorrect implementation of the method.
%
While the previous interactions highlighted how maintainers often favor readability over strict code deduplication, this case added a new dimension to our understanding: the \textit{cost-benefit} evaluation of proposed changes. A maintainer pointed out that the effort required to propagate such modifications after acceptance (i.e., during the upstreaming process) could outweigh the perceived benefits: ``\textit{[...] such patches might not worth it since the proposed improvement is very small (and questionable) while the upstreaming process still requires some effort.}''
From this feedback, we crafted our third takeaway:
%\vspace{-.5cm}
\begin{highlight-box}{Takeaway T3 (Integration Overhead)}
Maintainers may evaluate the trade-off between the impact of contributions
(not only deduplications) and the struggle to integrate them into the complex
Linux project.
\end{highlight-box}

@@

\begin{highlight-box}{Takeaway T5 (Performance)}
  Given the low-level development context of Linux and the IIO subsystem, which
  is usually related to devices with limited resources, we detected hints that
  duplication may be forgivable if it improves performance.
\hfill\\
\end{highlight-box}

@@

We derived four takeaways from \textbf{participant-as-observer observation}. These takeaways are accompanied by example quotes from the maintainers that originated during the review process~\footnote{It is paramount to state that we \textbf{are not affirming} that these quotes represent the complete positioning of the maintainers who wrote them on code duplication.}.

@@

Ethnographic studies, including full participant and participant-as-observer observations, provided a realistic view of the opportunities and challenges in removing code duplications detected by the tool. First-time Linux kernel contributors could effectively use the tool to identify duplications and submit patches.

While this research included one successfully merged patch by the authors, and 7 out of 12 patches sent by students were accepted and merged into Linux, a portion of the students' efforts encountered hurdles: 4 student groups had their patches rejected.

Nevertheless, our work suggests that maintainers do not always view code duplication as inherently unfavorable, especially when it improves clarity, performance, and more. This suggests a nuanced understanding of code quality within the Linux kernel community that can go against the DRY principle, where practicality and maintainability in specific contexts sometimes outweigh the general principle of eliminating all code duplication.

>>>>

\paragraph*{\textbf{Deduplication Option 1}} We ran the tool on the IIO subsystem to generate a list of duplicated function pairs, then curated it to highlight actionable cases for students. Specifically, we (1) kept only pairs within the same file, since cross-driver duplications could involve multiple maintainers or distinct interfaces; (2) removed very short duplications; and (3) manually ranked the remaining pairs, adding annotations to guide students toward entries likely to be accepted. In this option, students were tasked with selecting a recommended entry, devising a way to remove the duplication, and submitting their patches to the IIO mailing list for review.

\paragraph*{\textbf{Deduplication Option 2}} We offered an experience similar to our initial complete participant observation, where students managed the entire workflow: running the tool, analyzing results to identify potential deduplications, creating a patch, and submitting it to the driver's maintainers. This option provided freedom of choice but made the students' task more complex.

@@

It is important to stress that, \textbf{no matter the chosen approach}, we supported all groups by providing review cycles on their contributions before submission, to prepare students for the interactions with the maintainers and alleviate rough, but natural, newcomers' mistakes. After the submission, we helped students understand the feedback from maintainers and gave directions on how to move forward. This close, yet indirect, participation through intimate mentoring characterizes the second phase as our participant-as-observer observation.

For this second phase, our primary data source was the mailing list threads, where all development occurred (available in our Replication Package). Using blog posts, the groups documented their experiences submitting a patch to the Linux kernel~[omitted], and we also leveraged those to complement our observations. % (available in our Replication Package).
%TODO: Add blogs ao pacote de replicação

@@

Since we suffered a delay in the review process, we directly contacted a maintainer to understand why. From this conversation, we understand that one reason for the delay was that duplicated code enhances the independence of GPU driver code. This approach allows developers to make changes to a specific GPU without needing to test compatibility with others, which helps save a significant amount of time and effort. This example of hardware variations, also mentioned by Kapser and Godfrey~\cite{cloneharm}, leads us to the primary conclusion of our complete participant observation:
%
\begin{highlight-box}{Takeaway T0}
Driver developers do not necessarily view duplicated code negatively, which contradicts standard software engineering practices.
\end{highlight-box}

@@

By the end of the feedback cycle, the patch from Group 4 was accepted, but the final implementation differed substantially from the original without using macros. The change was primarily driven by a maintainer's feedback: \textit{``I think the old code is more readable than hiding the values in a macro even if it is duplicating a few lines of code.''}.
%
This response category (also from groups 1, 2, and 10) revealed a recurring concern among maintainers that went beyond functional correctness or redundancy reduction. Instead, it reflected a preference for code clarity and comprehension. Building on these consistent insights from multiple review interactions, we formulated our first takeaway:
%\vspace{-.5cm}
\begin{highlight-box}{Takeaway T1 (Readability)}
Maintainers prioritized readability over removing duplications in many different contexts. In their view, going against the \textbf{DRY principle} was worth it if it resulted in easier-to-read and understand code.
\end{highlight-box}

Still focusing on Group 1, maintainers considered the patch irrelevant. The reason was that Group 1 chose to apply the refactoring to a device containing a register that was both read-only and non-volatile. Thus, the technical debt selected by Group 1 was not merely a duplication but an incorrect implementation of the method.
%
While the previous interactions highlighted how maintainers often favor readability over strict code deduplication, this case added a new dimension to our understanding: the \textit{cost-benefit} evaluation of proposed changes. A maintainer pointed out that the effort required to propagate such modifications after acceptance (i.e., during the upstreaming process) could outweigh the perceived benefits: ``\textit{[...] such patches might not worth it since the proposed improvement is very small (and questionable) while the upstreaming process still requires some effort.}''
From this feedback, we crafted our second takeaway:
%\vspace{-.5cm}
\begin{highlight-box}{Takeaway T2 (Integration Overhead)}
Maintainers may evaluate the trade-off between the impact of contributions
(not only deduplications) and the struggle to integrate them into the complex
Linux project.
\end{highlight-box}

@@

\begin{highlight-box}{Takeaway T5 (Performance)}
  Given the low-level development context of Linux and the IIO subsystem, which
  is usually related to devices with limited resources, we detected hints that
  duplication may be forgivable if it improves performance.
\end{highlight-box}

@@

@@

This ethnographic study, combining complete participant and participant-as-observer observations, provided a realistic view of the opportunities and challenges in removing code duplications detected by the tool. First-time Linux kernel contributors could effectively use the tool to identify duplicated functions and submit meaningful patches.

Throughout this process, we observed that maintainers do not necessarily regard code duplication as inherently harmful. Instead, their decisions reflect a nuanced and context-dependent understanding of software quality. As evidenced by our takeaways\footnote{It is paramount to state that we \textbf{are not affirming} that these quotes represent the complete positioning of the maintainers who wrote them on code duplication.}, readability often outweighed strict adherence to the DRY principle (T1), while the effort required to propagate changes also influenced maintainers’ acceptance decisions (T2). Moreover, avoiding cognitive overhead when navigating the code (T3) and preserving performance in resource-constrained environments (T4) emerged as additional justifications for tolerating certain forms of duplication.

Overall, these findings reveal that, within the Linux kernel community, the evaluation of code duplication extends beyond technical redundancy. Maintainability, clarity, and practical trade-offs play a decisive role in shaping maintainers’ judgments, suggesting that DRY is sometimes less valuable than keeping the code understandable, efficient, and integrated smoothly into a large and evolving system.
